Clustering: A pipeline to analyse time-course gene expression data

### 2019-04-25 scripts/00_introduction.Rmd

---
title: "Introduction: a pipeline to analyse time-course gene expression data"
output: rmarkdown::html_vignette
---

@@

While there are many methods that have been proposed for discrete aspects of
time course data, the entire workflow for analysis of such data remains
difficult, particularly for long, developmental time series. Most methods
proposed for time course data are concerned with detecting genes that are
changing over time (differential expression analysis), examples being
\texttt{edge}\todo[Other methods to mention: impulse models, what else?
Citations!]. However, with long time course datasets, particularly in
developmental systems, a massive number of genes will show *some* change.
For example, in the \xxx system, over \xxx\% of genes are shown to be changing
over time. The task in these settings is often not to detect changes in genes,
but to categorize into biologically interpretable patterns the vast number of
changes discovered.

We present here a workflow for such an analysis that consists of \xxx main
parts (Figure \ref{fig:workflow}):

- Identification of genes that are differentially expressed
- Clustering of genes into distinct temporal patterns
- Biological interpretation of the clusters

**FIXME EP This is probably not best enumeration of steps, but need
something...**

This workflow represents an integration of both novel implementations of
previously established methods and new methodologies for the settings of
developmental time series. We provide the various steps of the workflow as
functions in a R package, \citettt{\xxx}, along with command-line tools for
running the entire workflow in a single command.

@@

This workflow is illustrated using data from a micro-array time-course
experiment, exposing mice to three different strains of influenza, and
collecting lung tissue during 14 time-points after infection (0, 3, 6, 9, 12,
18, 24, 30, 36, 48, 60 hours, then 3, 5, and 7 days later). The three strains
of influenza used in the study are (1) a low pathogenicity seasonal H1N1
influenza virus (A/Kawasaki/UTK4/2009 [H1N1]), a mildly pathogenic virus from
the 2009 pandemic season (A/California/04/2009 [H1N1]), and a highly
pathogenic H5N1 avian influenza virus (A/Vietnam/1203/2004 [H5N1]. Mice were
injected with 10^5 PFU of each virus. An adidtional 42 mice were injected with
a lower dose of the Vietnam avian influenza virus (10^3 PFU).

@@

>>>>

---
title: "Introduction: a pipeline to analyse time-course gene expression data"
output: rmarkdown::html_vignette
bibliography: bibliography.bib
---

@@

While there are many methods that have been proposed for discrete aspects of
time course data, the entire workflow for analysis of such data remains
difficult, particularly for long, developmental time series. Most methods
proposed for time course data are concerned with detecting genes that are
changing over time (differential expression analysis), examples being
`edge` [@storey:significance], functional component analysis based models
[@wu:more], time-course permutation tests [@park:statistical], XXX
[@sun:multiple]
**TODO: Other methods to mention: impulse models, what else?
Citations!**. However, with long time course datasets, particularly in
developmental systems, a massive number of genes will show *some* change.
For example, in the \xxx system, over \xxx\% of genes are shown to be changing
over time. The task in these settings is often not to detect changes in genes,
but to categorize into biologically interpretable patterns the vast number of
changes discovered.

We present here a workflow for such an analysis that consists of 4 main
parts (Figure \ref{fig:workflow}):

- Quality control and normalization;
- Identification of genes that are differentially expressed;
- Clustering of genes into distinct temporal patterns;
- Biological interpretation of the clusters.

This workflow represents an integration of both novel implementations of
previously established methods and new methodologies for the settings of
developmental time series. We provide the various steps of the workflow as
functions in a R package **FIXME**.

@@

This workflow is illustrated using data from a micro-array time-course
experiment, exposing mice to three different strains of influenza, and
collecting lung tissue during 14 time-points after infection (0, 3, 6, 9, 12,
18, 24, 30, 36, 48, 60 hours, then 3, 5, and 7 days later)
[@shoemaker:ultrasensitive]. The three strains of influenza used in the
study are (1) a low pathogenicity seasonal H1N1 influenza virus
(A/Kawasaki/UTK4/2009 [H1N1]), a mildly pathogenic virus from the 2009
pandemic season (A/California/04/2009 [H1N1]), and a highly pathogenic H5N1
avian influenza virus (A/Vietnam/1203/2004 [H5N1]. Mice were injected with
10^5 PFU of each virus. An adidtional 42 mice were injected with a lower dose
of the Vietnam avian influenza virus (10^3 PFU).

@@

## References

### 2019-04-25 scripts/03_clustering.Rmd

# Only assign labels to X% of the genes
max_score = quantile(moanin:::row_min(all_scores), c(percentage_genes_to_label))
genes_to_not_consider = moanin:::row_min(all_scores) >= max_score
labels = moanin:::row_argmin(all_scores)
labels[genes_to_not_consider] = NA
```

```{r}
table(kmeans_clusters$clusters)
```

@@

```{r results="hide", echo=FALSE}
# Save the clustering for the next step
kmeans_labels = as.data.frame(labels)
row.names(kmeans_labels) = row.names(y)
write.table(kmeans_labels,
	    ".results/clustering_labels.txt",
	    sep="\t")
```

>>>>

# Only assign labels to X% of the genes
max_score = quantile(moanin:::row_min(all_scores), c(percentage_genes_to_label))
genes_to_not_consider = moanin:::row_min(all_scores) >= max_score
labels = moanin:::row_argmin(all_scores)
labels[genes_to_not_consider] = NA
names(labels) = row.names(data)
```

Summary of number of genes assigned with original scoring, and "better"
scoring.

```{r}
table(kmeans_clusters$clusters)
barplot(table(kmeans_clusters$clusters), col="black")
```

```{r}
table(labels)
barplot(table(labels), col="black")
```

Just a hack, cause I didn't update the code
```{r}
kmeans_clusters$clusters = labels[!is.na(labels)]
```

```{r results="hide", echo=FALSE}
# Save the clustering for the next step
kmeans_labels = as.data.frame(labels[!is.na(labels)])
write.table(kmeans_labels,
	    ".results/clustering_labels.txt",
	    sep="\t")
```

Now let's visualize the centroids.

@@
